---
title: "Project 8"
author: "biRds"
date: "2023-12-03"
output: html_document
bibliography: "BIOL3140.bib"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(rgbif)
library(tidyverse)
library(MuMIn)
library(rnoaa)
library(data.table)
library(ggmap) 
library(usmap) 
library(magick)
library(cowplot) 
library(lme4)
library(car) 
library(data.table) 
library(knitr)
library(kableExtra)
```

## Introduction 
Trans-Gulf migrants (TGMs) are a subset of birds migrating from Central and South America to Texas and Florida, traversing the Gulf of Mexico non-stop. Their arrival dates, a focal point for researchers, may influence breeding success, with debates on whether early arrivals or those benefiting from increased food availability fare better, as discussed by @Newton2008 and @Smith2005. Climate change, as highlighted by @Both2006, poses a threat, potentially reducing TGM populations due to mismatches with altered spring temperatures. @Cook2021's studies reveal cold weather and high population density negatively impact survival rates, exacerbating challenges for TGMs amid climate-induced fluctuations.

In response, our project integrates eBird, a citizen science initiative by the Cornell Lab of Ornithology and the National Audubon Society, with meteorological data to assess the impact of local weather on TGM arrivals in Massachusetts. By leveraging this unique combination, our goal is to deepen insights into how TGMs adapt to changing environmental conditions, contributing significantly to avian biology research and aiding conservation efforts.

## Methods 
Our research focused on five distinct trans-Gulf migrant (TGM) bird species representing various families: Myiarchus crinitus, Icterus galbula, Pheucticus ludovicianus, Coccyzus americanus, and Setophaga caerulescens. Leveraging eBird data from Massachusetts and the rgbif package's occ_data() function, we obtained occurrence information for these species. The eBird dataset allowed us to determine Julian days for each species, facilitating an exploration of temperature and wind influences along their migration route. We examined three key locations—Mobile, AL (Gulf initiation), Charlotte, NC (midpoint), and Boston, MA (final destination)—with NOAA weather data aiding in plotting arrival proportions and predicted times for each species. Temporal variations in Julian days over the years were also visualized to assess climate impacts on arrival dates.

Wind direction, calculated from weather data, was integrated with species data. Using the frollmean() function, we determined the mean of weather variables for both 5 and 10 days preceding arrival. Subsequently, a Linear Mixed-effect Model was constructed with the lme package, considering both single-day and two-week average weather datasets. An Anova test identified the best-fit model, and model testing was executed with the dredge() function. This comprehensive approach allowed us to elucidate the intricate relationship between weather parameters and the arrival dynamics of selected TGM bird species along their migratory journey.


## Results 
```{r, echo=FALSE, message =FALSE}
raven <- occ_data(scientificName = "Corvus corax", stateProvince="Maine", limit=200,year=2018)

ME<- map_data('state', 'maine')

raven.p <- ggplot(ME, aes(long,lat,group=subregion) )+
  geom_polygon(colour = "gray",fill="gray90")+geom_point(data=raven[[2]],aes(x=decimalLongitude,y=decimalLatitude,size=individualCount),alpha=0.3,inherit.aes = F)+ coord_quickmap()+theme_void()

raven.p2 <- ggdraw() +
  draw_image("https://www.allaboutbirds.org/guide/assets/photo/63739491-480px.jpg",scale = 0.3,halign=0,valign=1) +
  draw_plot(raven.p)
print(raven.p2)

lapply(raven,head)

species <- c("Myiarchus crinitus","Icterus galbula","Pheucticus ludovicianus","Coccyzus americanus","Setophaga caerulescens")

y <- paste0("2000",",","2019")
m <- paste0("4",",","5")

dat.l <-list()

for(s in species){
  
n.obs <-  occ_data(scientificName = s,year=y,month=m,limit=0,country="US",basisOfRecord = "HUMAN_OBSERVATION",stateProvince="Massachusetts")$meta$count 

print(n.obs)


dat.l[[paste0(s)]] <- occ_data(scientificName = s,year=y,month=m,
                               limit=n.obs,country="US",
                               basisOfRecord = "HUMAN_OBSERVATION",
                               stateProvince="Massachusetts")[[2]]}

dat <- rbindlist(dat.l,fill=T)

head(dat)
```
The results above were from a for loop  that gets the occurence data in in MA during April and May in the years 2000-2019.  

```{r, echo=FALSE, message=FALSE}
saveRDS(data,"massbird.data.RDS")

dat%>%
  group_by(year,species)%>%
  summarise(count=sum(individualCount,na.rm = T))%>%
  ggplot(aes(x=year,y=count,col=species))+geom_point()

options(noaakey = "MSwYPqBXDAEWVsfKiGtHDmZODJaXABNT")
sts <- c(
  "GHCND:USW00013894", #Mobible, AL 2k away about 10 days away @200 km/day
  "GHCND:USW00013881", #Charlotte, NC 1000 km away about 6 days away @200 km/day
  "GHCND:USW00014739" #Boston
)
bos <- ncdc_stations(stationid = "GHCND:USW00014739")
print(bos)

Sys.setenv(NOAA_KEY = "MSwYPqBXDAEWVsfKiGtHDmZODJaXABNT")

sta.d <- bind_rows( 
  lapply(sts,function(x) ncdc_stations(stationid = x)$data ) 
  )%>%
  mutate(usmap_transform(.,input_names = c("longitude","latitude"),output_names = c("longitude.1", "latitude.1")))%>% 
  mutate(name=str_sub(name, -5,-4))%>%
  mutate(migr.day=c(10,5,0))%>%
  separate(id,into = c("station.type","id"))%>%
        print()
plot_usmap(
  include = c(.northeast_region,.south_region,.east_north_central)
)+geom_point(data=sta.d,aes(x=longitude.1,y=latitude.1,col=name),size=5)+geom_label(data=sta.d,aes(x=longitude.1,y=latitude.1,col=name,label=name),size=5,nudge_x = 1e6*0.25)+theme(legend.position = "none")

weather.d <- meteo_pull_monitors(sta.d$id,date_min = "2000-01-01")

mc<- dat%>%
  group_by(year)%>%
  mutate(date=as.Date(paste0(year,"-",month,"-",day)),
         j.day=julian(date,origin=as.Date(paste0(unique(year),"-01-01")))
  )%>%
  group_by(species,year,j.day,date)%>%
  summarise(day.tot=sum(individualCount,na.rm=T))%>%
  group_by(species,year)%>%
  mutate(prop=cumsum(day.tot/sum(day.tot,na.rm = T)))%>%
  filter(year>1999)

mc%>%
  ggplot(aes(j.day,prop))+geom_point()+facet_wrap(year~.)

mc.pred <- mc%>%
  group_by(year)%>%
 summarize(
   pred=predict(nls(prop~SSlogis(j.day,Asym, xmid, scal)),newdata=data.frame(j.day=min(j.day):max(j.day))),#predict the logistic curve for each species
   j.day=min(j.day):max(j.day),
  )%>%
  left_join(mc%>%dplyr::select(j.day,date))

mc%>%
  ggplot(aes(j.day,prop))+geom_point(aes=0.3)+geom_line(data=mc.pred,aes(x=j.day,y=pred),col="blue",size=2)+facet_wrap(year~.) 

mc.arrive.date <-mc.pred%>%
  group_by(year)%>%
  filter(j.day==j.day[which.min(abs(pred-0.25))])

mc.arrive.date%>%
  ggplot(aes(year,j.day))+geom_point()

weather.d <- weather.d%>%
  mutate(year=as.integer(str_sub(date,1,4)), 
         date=as.Date(date))%>%
  group_by(year)%>% 
 mutate(j.day=julian(date,origin=as.Date(paste0(unique(year),"-01-01"))), 
  date2=date,
  wdir.rad=(180-abs(wdf2-180))*pi/180, 
  wvec=cos(wdir.rad)*-1*awnd) %>%
  dplyr::select(id,year,date2,j.day,tmin,tmax,wvec)%>% 
  left_join(sta.d%>%select(id,name,migr.day))%>% 
  mutate(j.day=j.day+migr.day)
mc.arr.weath <- mc.arrive.date%>%
  left_join(weather.d)%>%
  left_join(mc%>%dplyr::select(year,date,j.day))

weather.wk <-weather.d %>% 
  group_by(year,name) %>% 
  mutate(wk.tmin = frollmean(tmin, n=14,align="right"),
         wk.tmax = frollmean(tmax, n=14,align="right"),
         wk.wvec = frollmean(wvec, n=14,align="right")
         )%>%
  dplyr::select(j.day,date2,name,wk.tmin,wk.tmax,wk.wvec)

mc.arr.weath2 <- mc.arrive.date%>%
  left_join(weather.wk)

mc.lmer <- lmer(j.day~tmin*tmax*wvec+(1|name),mc.arr.weath,na.action = "na.fail")

Anova(mc.lmer)

mc.lmer2 <- lmer(j.day~wk.tmin*wk.tmax*wk.wvec+(1|name),mc.arr.weath2,na.action = "na.fail")
Anova(mc.lmer2) 

mc.arr.aic <- dredge(mc.lmer2,fixed = c("wk.tmin","wk.tmax","wk.wvec"),)

mc.kb <- kable(mc.arr.aic[1:4,],caption = "Fit values for nested models of the most complicated lme model")

kable_styling(mc.kb)

best.lmer <-  lmer(j.day~wk.tmin+wk.tmax+wk.wvec+(1|name),mc.arr.weath2,na.action = "na.fail")

Anova(best.lmer)
```


## Discussion 
The best-fit model includes minimum and maximum temperature and wind vector. According to the ANOVA test, all three aforementioned factors are significant predictors. 

In regard to wind as a migratory factor, many migratory birds migrate at night utilizing landmarks in the night sky and the absence of daytime thermals. In addition, winds tend to be more mild or tailwinds. However, climate change affects multiple factors in our environment including atmospheric circulation. There is evidence that some North American migratory birds will account for wind patterns and adjust their heading to avoid drift or fly at certain altitudes to take advantage of head- and tailwinds [@LaSorte2018]. 

Maximum temperatures appear to have a more significant effect on arrival dates than minimum temperatures. In regard to this, a study conducted in 2012 examining the relationship between arrival date in North American migratory birds and warming temperatures showed that birds arrived earlier as temperature increased [@Hurlbert2012]. More specifically, species arrived 0.8 days earlier for every 1°C increase in temperature. In addition to this, they also saw species that would arrive 3-6 days earlier per 1°C of warming. Also in regard to local temperature changes, studies have shown differences in arrival date shifts between short-distance and long-distance migrants.

While some species appear to be arriving earlier in correlation with rising temperatures, others are not. A 2015 study focusing on migratory birds in the northern Great Plains reported both significant changes in early and late arrival dates depending on the species [@Travers2015]. In all, the growing seasons for North American migratory birds are generally increasing. While both this experimental study and others suggest that species, including trans-Gulf migrants, are adjusting their arrival dates in relation to this change, the rate at which climate change is occurring may be too rapid for species to effectively adapt to.

## Author Contributions
Alexandra Hoogendijk: Results

Ava Skogstrom: Discussion, Bibliography 

Tara Gerjarusak	

Susanne Hahs: Introduction, Methods, Bibliography


## References
